Optimal updating factor in adaptive flat-distribution-sampling simulations

### 2017-06-17 doc/invt.tex
invt.tex: various corrections, remove Appendix B and stabilization

In this study,
we present a method of computing
the optimal schedule
for an adaptive FDS method
of a general updating scheme,
including the single-bin and Gaussian one
used in the WL algorithm and metadynamics.
%
Our method maps the optimization problem to a mechanical one,
in which the schedule plays the role of the velocity of
a free particle with a position-dependent mass,
and the final error becomes the action.
%
In this way, the optimal schedule
can be deduced from the equation of motion
of the particle that minimizes the action.
%
The resulting optimal schedule
depends on the updating window function
through the mass function.
%
While the optimal schedule in the single-bin case
is given by the known inverse-time formula,
that for Gaussian case is more complex
and sensitive to the simulation length
and the width of the Gaussian window.
%
%In either case, however,
%Ideally,
%the optimal initial updating magnitude
%can be set to
%half of the previous equilibrium value,
%which amounts to a shift of the origin of time
%in the single-bin case.
%
Further, we will show that
the single-bin updating scheme
is optimal for asymptotic convergence,
and it can be generalized
to a class of asymptotically optimal
bandpass updating schemes.
%
The latter, however,
allow one to selectively flatten only
a few long-range histogram modes,
and to avoid introducing local noise
into the bias potential during the updating process.
%
The optimal schedule of bandpass updating schemes
is also the above origin-shifted inverse-time schedule.
%
Interestingly, several previous FDS algorithms\cite{
  langfeld2012, pellegrini2014,
  neuhaus2006, *neuhaus2007}
can be cast as special cases of bandpass updating schemes.

@@

We develop the theory
in the following order.
%
In Sec. \ref{sec:background},
we first review the basics of FDS
and some known aspects of the optimal schedule.
%
Then, in Sec. \ref{sec:single-bin},
we illustrate the method of
computing the optimal schedule
in the simple case of the
single-bin updating scheme
used by the WL algorithm,
by proving the optimality
of the known inverse-time formula.
%
The above two introductory sections help define our notations,
and an expert reader may wish to skip them.
%
Next, we extend the method to the general case
of a multiple-bin updating scheme
that encompasses the Gaussian scheme used in metadynamics
in Sec. \ref{sec:multiple-bin},
and present a general procedure of computing the optimal schedule
in Sec. \ref{sec:procedure}.
%
Finally, we compare different updating schemes
and show that the single-bin scheme
and a class of generalized bandpass schemes
are optimal for convergence
in the long time or extensive sampling limit
in Sec. \ref{sec:cmpschemes}.

@@

%For the single-bin scheme
%used in WL algorithm,
A more effective way
of controlling the schedule, $\alpha(t)$,
in late stages
is to follow the formula
%
\begin{equation}
  \alpha(t) = \frac{1}{t},
  \label{eq:alpha_invt}
\end{equation}
%
where $t$,
referred to as the ``time'' below,
is the number of updating steps
from the beginning of the simulation\cite{
belardinelli2007, *belardinelli2007jcp, *belardinelli2008, *belardinelli2016,
morozov2007, zhou2008,
komura2012, *caparica2012, *caparica2014}.
%
This can be derived from a general result\cite{
  robbins1951, pellegrini2014}
and an intuitive explanation\cite{
  marsili2006, barducci2008}
comes from the interpretation of
the bias potential %under the WL updating scheme, Eq. \eqref{eq:wl_update},
as a runtime average computed from all data points collected so far,
so that the addition of a new data point only
affects the average by its due weight,
the inverse of the current sample size, $t$
(cf. Appendix \ref{sec:equilerr}).
%
We may also derive it
from optimally planning the simulation lengths in the stages
and the reduction factor in the WL setup (cf. Appendix \ref{sec:wltoinvt}).

@@

Second, the bias potential following Eq. \eqref{eq:wl_update}
generally grows over time.
%
Since a uniform increase of $u_i(t)$, or, equivalently, that of $v_i(t)$,
does not affect the resulting distribution, $\pi_i(t)$,
the real deviation of the bias potential
depends only on the difference between $v_i(t)$
and a baseline value, $\bav{v}(t)$,
for average growth\cite{
dama2014}.
%
We define a bin-average-deducted bias potential
as\footnote{Alternatively,
we may define $v_{*i}(t)$ such that
$\rho_i \, e^{-v_{*i}(t)}$ is a normalized distribution,
i.e.,
$$
v_{*i}(t) \to -\ln\left(
  \frac{ (p_i/\rho_i) \, e^{ -u_i(t) } }
  { \sum_{j=1}^n p_j \, e^{ -u_j(t) } }
\right).
$$
One can readily show that the two definitions are equivalent
for small $v_{*i}$.}
%
\begin{equation}
  v_{*i}(t) \equiv v_i(t) - \bav{v}(t)
  ,
\label{eq:x_def}
\end{equation}
%
where the bin average, $\bav{v}(t)$, is defined as
$\bav{v}(t) \equiv \sum_{i=1}^n \rho_i \, v_i(t)$.
%
%\begin{equation}
%  \bav{v}(t)
%  \equiv \sum_{i = 1}^n \rho_i \, v_i(t)
%  .
%  \label{eq:vav_def}
%\end{equation}
%
Below we will use the notations, $x_{*i}$ and $\bav{x}$,
for an arbitrary variable, $x_i$.
%
The total error is given by
%
\begin{equation}
  \Err(t)
  =
  \sum_{i = 1}^n \rho_i \,
  \left\langle v_{*i}^2(t) \right\rangle
  .
\label{eq:error_def}
\end{equation}

@@

From Eqs.
\eqref{eq:vt_diffeq}-\eqref{eq:sh_ave},
we get % a set of decoupled equations
%
\begin{equation}
  \dot v_{*i}(t)
  =
  \alpha(t) \, \left[ \langle f_{*i}(t) \rangle + \zeta_{*i}(t) \right]
  \approx
  -\alpha(t) \, \left[ v_{*i}(t) - \zeta_{*i}(t) \right]
  ,
  \notag
  %\label{eq:dxdt_singlebin}
\end{equation}
%
and the formal solution is
%
\begin{equation}
  v_{*i}(T)
=
  v_{*i}(0) \, e^{-q(T)}
+
\int_0^T
  \dot y\bigl( q(t) \bigr) \, \zeta_{*i}(t) \, dt,
\notag
%\label{eq:xt_solution}
\end{equation}
%
where
%
$
q(t) \equiv \int_0^t \alpha(t') \, dt',
$
%
and
%
\begin{align}
y(q')
&\equiv
e^{q' - q(T)}.
\label{eq:y_def}
\end{align}

@@

At the end of period $T$,
the total error defined in Eq. \eqref{eq:error_def}
is given by
%
\begin{align}
  \Err(T)
  =
  \Err(0) \, e^{-2 \, q(T)}
  +
  \Gamma \,
  \int_0^T
    {\dot y}^2\bigl( q(t) \bigr) \,
    \, dt
  ,
  \label{eq:ET_average}
\end{align}
where $\Gamma \equiv \sum_{i=1}^n \rho_i \, G_i$.
%\begin{equation}
%  \Gamma \equiv \sum_{i=1}^n \rho_i \, G_i.
%  \label{eq:Gamma_from_G}
%\end{equation}

The error depends implicitly on the curve, $\alpha(t)$,
or, equivalently, the curve, $q(t)$,
via $v_{*i}(T)$, and
our aim is to find the $q(t)$
that minimizes this error.
%
We now fix the endpoint value, $q(T)$,
and vary the curve, $q(t)$ ($0 < t < T$).
%
This is equivalent to varying the curve $y(q(t))$
with fixed endpoints,
$y\bigl( q(0) \bigr)  = e^{- q(T)}$
and
$y\bigl( q(T) \bigr) = 1$.
%
Note also that the first term on the right-hand side
of Eq. \eqref{eq:ET_average} is fixed during
the process,
while the second term is the action of a free particle.
Thus, the velocity of $y(q(t))$ should be a constant:
%
\begin{equation}
  \dot y\bigl( q(t) \bigr) = c
  .
\label{eq:dydt_const}
\end{equation}
%
Then, by using Eq. \eqref{eq:y_def}, we have
%
\begin{equation}
  e^{ q(t) - q(T) }
  =
  c \, (t + t_0)
  =
  \frac{ t + t_0 }
       { T + t_0 }
  ,
\label{eq:expqt}
\end{equation}
where $t_0$ is some constant,
and $c$ has been determined
from the $t = T$ case
in the second step.
%
Taking the logarithm,
and differentiating both sides
with respect to $t$,
we get the inverse-time schedule
%
\begin{equation}
  \alpha(t) = \frac{ 1 }{ t + t_0 }
  ,
\label{eq:alpha_invt1}
\end{equation}
%
and Eq. \eqref{eq:alpha_invt}
is the special case of $t_0 = 0$.
%

We now determine the optimal value of $q(T)$,
which is fixed in the above process.
%
This is equivalent to choosing a value of $t_0$
\big[because $q(T) = \ln\bigl[1 + (T/t_0)\bigr]$,
from the $t = 0$ case of Eq. \eqref{eq:expqt}\big].
%
By using Eq. \eqref{eq:alpha_invt} in Eq. \eqref{eq:ET_average},
we find the final total error to be
%
$\Err(T) = [ \Err(0) \, t_0^2 + \Gamma \, T ]/(t_0 + T)^2$,
%
which is minimal at
%
\begin{equation}
  t_0 = \frac{ \Gamma } { \Err(0) }
  ,
  \notag
  %\label{eq:t0_singlebin}
\end{equation}
%
and the minimum is
%
\begin{equation}
  \Err(T)
  =
  \frac{ \Gamma } { T + t_0 }
  .
  \label{eq:Emin_singlebin}
\end{equation}
%
This error is roughly equal to
the average histogram fluctuation\cite{zhou2005, zhou2008}
(see also Appendix \ref{sec:hfluc}).

@@

We now turn to a general multiple-bin updating scheme.
%
By definition, Eq. \eqref{eq:mbin_update},
a visit to bin $j$ in this case results in updates of the bias potential
not only at bin $j$, but also at a few neighboring bins $i$'s.
%
Similar to Eq. \eqref{eq:ut_diffeq},
we can approximate the updating scheme
as a differential equation,
%
\begin{equation}
  \dot u_i(t)
  \approx
  \alpha(t) \,
  \sum_{j=1}^n w_{ij} \, \frac{ h_j(t) } { \rho_j }
  .
  \label{eq:ut_diffeq_mbin}
\end{equation}
%
The optimization procedure is similar to
the single-bin case.
%
The key step comes from the
projection of the bias potential to the $n$ eigenmodes
of the updating matrix, $\mathbf w$,
formed by the relative magnitudes, $w_{ij}$'s.

\subsubsection{\label{sec:updating-matrix}
Updating matrix}

The matrix, $\mathbf w$, is subject to three conditions.
%
First, we have a fixed-point condition\cite{bussi2006, dama2014}.
%
To sample the desired distribution,
$\pmb\rho = (\rho_1, \dots, \rho_n)$,
the growth rate of the bias potential,
${\dot u}_i(t)$
in Eq. \eqref{eq:ut_diffeq_mbin},
should be a constant of $i$
if $h_j(t)$ were the same as $\rho_j$,
i.e.
$\sum_{j=1}^n w_{ij}$ should be a constant
to allow the bias potential to grow uniformly
in the asymptotic regime.
%
By a proper overall scaling of $\mathbf w$,
we may write this condition as
%
\begin{equation}
  \sum_{j = 1}^n w_{ij} = 1
  .
\label{eq:w_sumj}
\end{equation}
%
In other words, $(1, \dots, 1)^T$
is a right eigenvector of $\mathbf w$
with eigenvalue $1$.
%
Thus, the transpose $\mathbf w^T$
resembles a transition matrix,
except that some elements can be negative
in our cases.
%
Below we will take advantage of the analogy
and borrow techniques used
in studying transition matrices\cite{vankampen}
(we will only use properties that are unaffected
by the admission of negative elements into $\mathbf w$).

@@

%It is worth pointing out
Finally, there is a stability condition.
%
In a \emph{stable} updating scheme,
no eigenmode grows indefinitely,
which means that, as we will show, no eigenvalue, $\lambda_k$,
can be negative for $k \ge 1$.

\subsubsection{Eigenmode decomposition}

We can further define
a generalized Fourier transform, $\mathcal{F}$,
for variable, $A_i$,
%
\begin{equation}
  {\tilde A}_k
  \equiv \mathcal{F}[A_i]_k
  \equiv \sum_{i = 1}^n \phi_{ki} \, A_i
  ,
  \notag
  %\label{eq:gft_def}
\end{equation}
%
and the inverse transform is
%
\begin{equation}
  A_i =
  \frac{1}{\rho_i} \sum_{k = 1}^n \phi_{ki} \, \tilde{A}_k
  .
  \notag
  %\label{eq:gft_inv}
\end{equation}
%
Using this definition,
we can rewrite Eq. \eqref{eq:ortho1} as
$\mathcal F[1]_k = \delta_{k0}$,
%
and Eq. \eqref{eq:eigenmode0} means that
the bin average of any quantity, $x_i$,
is the first mode of the Fourier transform:
$\bav{x}= \mathcal F[x_i]_0 = \tilde x_0$.
%
So
\begin{equation}
  \tilde x_{*k}
  = \tilde x_k - \bav{x} \, \delta_{k0}
  =
  \begin{dcases}
    0           & k = 0, \\
    \tilde x_k  & k > 0,
  \end{dcases}
  \label{eq:xtstar}
\end{equation}
%
which means that the only effect
of the bin-average deduction on $\tilde x_k$ is to
annihilate the first mode, $\tilde x_0$.

We can now use Eq. \eqref{eq:eig_w} to diagonalize
the multiple-bin updating scheme, Eq. \eqref{eq:mbin_update},
as
%
$$
{\tilde v}_k(t + 1) =
{\tilde v}_k(t) + \alpha(t) \, \lambda_k \,
{\tilde f}_k(t)
,
$$
or in continuous time, we have, for $k > 0$,
%
\begin{equation}
  \dot{\tilde v}_k(t)
  =
  \alpha(t) \, \lambda_k \, {\tilde f}_k(t)
  \approx
  -\alpha(t) \, \lambda_k \,
  \bigl[ {\tilde v}_k(t) - {\tilde \zeta}_k(t) \bigr]
  ,
  \label{eq:vt_diffeq_mbin}
\end{equation}
%
where we have removed the bin average,
and used the transformed version of Eq. \eqref{eq:sh_ave}
[but the subscript ``$_*$'' is dropped by Eq. \eqref{eq:xtstar}].
In the second step, we have used the linear approximation
as in Sec. \ref{sec:sbin_diffeq}.
%
Equation \eqref{eq:vt_diffeq_mbin}
shows that the rate of error reduction
is now mode-dependent,
and the minimization of the final error
involves a compromise of the individual modes.

@@

The error defined in Eq. \eqref{eq:error_def}
can be rewritten in terms of the eigenmodes of $\mathbf w$,
%
\begin{align}
  \Err(T)
  =
  \sum_{i, j = 1}^n \sum_{k=0}^{n-1}
   \phi_{ki} \, \phi_{kj}
    \left\langle v_{*i}(T) \, v_{*j}(T) \right\rangle
  =
  \sum_{k = 1}^{n - 1}
    \left\langle
      {\tilde v}_k^2(T)
    \right\rangle
  ,
\notag
\end{align}
%
where we have used
Eq. \eqref{eq:eig_orthonormal_cols}
and drop the first term
by using Eq. \eqref{eq:xtstar}
in the second step.

>>>>

In this study,
we present a method of computing
the optimal schedule
for an adaptive FDS method
of a general updating scheme,
including the single-bin and Gaussian one
used in the WL algorithm and metadynamics.
%
Our method maps the optimization problem to a mechanical one,
in which the schedule plays the role of the velocity of
a free particle with a position-dependent mass,
and the final error becomes the action.
%
In this way, the optimal schedule
can be deduced from the equation of motion
of the particle that minimizes the action.
%
The resulting optimal schedule
depends on the updating window function
through the mass function.
%
While the optimal schedule in the single-bin case
is given by the known inverse-time formula,
that for Gaussian case is more complex
and sensitive to the simulation length
and the width of the Gaussian window.
%
%In either case, however,
%Ideally,
%the optimal initial updating magnitude
%can be set to
%half of the previous equilibrium value,
%which amounts to a shift of the origin of time
%in the single-bin case.
%
Further, we will show that
the single-bin updating scheme
is optimal for asymptotic convergence,
and it can be generalized
to a class of asymptotically optimal
bandpass updating schemes.
%
The latter, however,
allow one to selectively flatten only
a few long-range histogram modes,
and to avoid introducing local noise
into the bias potential during the updating process.
%
The optimal schedule of bandpass updating schemes
is also the inverse-time schedule.
%
Interestingly, several previous FDS algorithms\cite{
  langfeld2012, pellegrini2014,
  neuhaus2006, *neuhaus2007}
can be cast as special cases of bandpass updating schemes.

@@

We develop the theory
in the following order.
%
In Sec. \ref{sec:background},
we first review the basics of FDS
and some known aspects of the optimal schedule.
%
Then, in Sec. \ref{sec:single-bin},
we illustrate the method of
computing the optimal schedule
in the simple case of the
single-bin updating scheme
used by the WL algorithm,
by proving the optimality
of the known inverse-time formula.
%
%The above two introductory sections help define our notations,
%and an expert reader may wish to skip them.
%
Next, we extend the method to the general case
of a multiple-bin updating scheme
that encompasses the Gaussian scheme used in metadynamics
in Sec. \ref{sec:multiple-bin}.
%
Finally, we compare different updating schemes
and show that the single-bin scheme
and a class of generalized bandpass schemes
are optimal for convergence
in the long time limit
in Sec. \ref{sec:cmpschemes}.

@@

%For the single-bin scheme
%used in WL algorithm,
A more effective way
of controlling the schedule, $\alpha(t)$,
in late stages
is to follow the formula
%
\begin{equation}
  \alpha(t) = \frac{1}{t},
  \label{eq:alpha_invt}
\end{equation}
%
where $t$,
referred to as the ``time'' below,
is the number of updating steps
from the beginning of the simulation\cite{
belardinelli2007, *belardinelli2007jcp, *belardinelli2008, *belardinelli2016,
morozov2007, zhou2008,
komura2012, *caparica2012, *caparica2014}.
%
This can be derived from a general result\cite{
  robbins1951, pellegrini2014}
and an intuitive explanation\cite{
  marsili2006, barducci2008}
comes from the interpretation of
the bias potential %under the WL updating scheme, Eq. \eqref{eq:wl_update},
as a runtime average computed from all data points collected so far,
so that the addition of a new data point only
affects the average by its due weight,
the inverse of the current sample size, $t$
(cf. Appendix \ref{sec:equilerr}).
%
%We may also derive it
%from optimally planning the simulation lengths in the stages
%and the reduction factor in the WL setup (cf. Appendix \ref{sec:wltoinvt}).

@@

Second, the bias potential following Eq. \eqref{eq:wl_update}
generally grows over time.
%
Since a uniform increase of $u_i(t)$, or, equivalently, that of $v_i(t)$,
does not affect the resulting distribution, $\pi_i(t)$,
the real deviation of the bias potential
depends only on the difference between $v_i(t)$
and a baseline value, $\bav{v}(t)$,
for average growth\cite{
dama2014}.
%
We define a bin-average-deducted bias potential
as\footnote{Alternatively,
we may define $v_{*i}(t)$ such that
$\rho_i \, e^{-v_{*i}(t)}$ is a normalized distribution,
i.e.,
$$
v_{*i}(t) \to -\ln\left(
  \frac{ (p_i/\rho_i) \, e^{ -u_i(t) } }
  { \sum_{j=1}^n p_j \, e^{ -u_j(t) } }
\right).
$$
One can readily show that the two definitions are equivalent
for small $v_{*i}$.}
%
\begin{equation}
  v_{*i}(t) \equiv v_i(t) - \bav{v}(t)
  ,
\label{eq:x_def}
\end{equation}
%
where we have defined the bin average,
$\bav{v}(t) \equiv \sum_{i=1}^n \rho_i \, v_i(t)$.
%
%\begin{equation}
%  \bav{v}(t)
%  \equiv \sum_{i = 1}^n \rho_i \, v_i(t)
%  .
%  \label{eq:vav_def}
%\end{equation}
%
Below we will use the notations, $x_{*i}$ and $\bav{x}$,
for a general variable, $x_i$.
%
The total error is given by
%
\begin{equation}
  \Err(t)
  =
  \sum_{i = 1}^n \rho_i \,
  \left\langle v_{*i}^2(t) \right\rangle
  .
\label{eq:error_def}
\end{equation}

@@

From Eqs.
\eqref{eq:vt_diffeq}-\eqref{eq:sh_ave},
we get % a set of decoupled equations
%
\begin{equation}
  \dot v_{*i}(t)
  =
  \alpha(t) \, \left[ \langle f_{*i}(t) \rangle + \zeta_{*i}(t) \right]
  \approx
  -\alpha(t) \, \left[ v_{*i}(t) - \zeta_{*i}(t) \right]
  ,
  \notag
  %\label{eq:dxdt_singlebin}
\end{equation}
%
and the formal solution is
%
\begin{equation}
  v_{*i}(T)
=
  v_{*i}(0) \, e^{-q(T)}
+
\int_0^T
  \dot \vartheta\bigl( q(t) \bigr) \, \zeta_{*i}(t) \, dt,
\notag
%\label{eq:xt_solution}
\end{equation}
%
where
%
$
q(t) \equiv \int_0^t \alpha(t') \, dt',
$
%
and
%
\begin{align}
\vartheta(q')
&\equiv
e^{q' - q(T)}.
\label{eq:theta_def}
\end{align}

@@

At the end of period $T$,
the total error defined in Eq. \eqref{eq:error_def}
is a sum of a residual error (from the decay of the initial error)
and an asymptotic error:
%
\begin{align}
  \Err(T)
  =
  \Err_R(T) + \Err_A(T)
  ,
  \label{eq:error_tot}
\end{align}
%
where
\begin{align}
  \Err_R(T)
  &= \Err(0) \, e^{-2 \, q(T)}
  ,
  \label{eq:ER_sbin}
  \\
  \Err_A(T)
  &= \Gamma \, \int_0^T \dot \vartheta^2\bigl( q(t) \bigr) \, dt
  ,
  \label{eq:EA_sbin}
\end{align}
with
$\Gamma \equiv \sum_{i=1}^n \rho_i \, G_i$.
%\begin{equation}
%  \Gamma \equiv \sum_{i=1}^n \rho_i \, G_i.
%  \label{eq:Gamma_from_G}
%\end{equation}

The error depends implicitly on the curve, $\alpha(t)$,
or, equivalently, the curve, $q(t)$,
via $v_{*i}(T)$.
%
If $\alpha(t)$ is a constant, $a_0$,
so that $q(t) = a_0 \, t$,
then after a long period,
the error approach an equilibrium
%
\begin{equation}
  \Err_\mathrm{equil.}
  = \Err_A = \Gamma \int_0^\infty e^{-2\,q'} \, a_0 \, d q'
  = \frac{1}{2} a_0 \Gamma
  .
  \label{eq:error_equil_sbin}
\end{equation}
%
Our aim is to find the $q(t)$
that minimizes this error.
%
We now fix the endpoint value, $q(T)$,
and vary the curve, $q(t)$ ($0 < t < T$).
%
This is equivalent to varying the curve $\vartheta(q(t))$
with fixed endpoints,
$\vartheta\bigl( q(0) \bigr)  = e^{- q(T)}$
and
$\vartheta\bigl( q(T) \bigr) = 1$.
%
Note also that the residual error
is fixed during the process,
while the asymptotic error is the action of a free particle.
Thus, the velocity of $\vartheta(q(t))$ should be a constant:
%
\begin{equation}
  \dot \vartheta\bigl( q(t) \bigr) = c
  .
\label{eq:dthetadt_const}
\end{equation}
%
Then, by using Eq. \eqref{eq:theta_def}, we have
%
\begin{equation}
  e^{ q(t) - q(T) }
  =
  c \, (t + t_0)
  =
  \frac{ t + t_0 }
       { T + t_0 }
  ,
\label{eq:expqt}
\end{equation}
%
where $t_0$ is some constant,
and $c$ has been determined
from the $t = T$ case
in the second step.
%
Taking the logarithm,
and differentiating both sides
with respect to $t$,
we get the inverse-time schedule
%
\begin{equation}
  \alpha(t) = \frac{ 1 }{ t + t_0 }
  ,
\label{eq:alpha_invt1}
\end{equation}
%
and Eq. \eqref{eq:alpha_invt}
is the special case of $t_0 = 0$.
%

We now determine the optimal value of $q(T)$,
which is fixed in the above.
%
This is equivalent to choosing a value of $t_0$
\big[because $q(T) = \ln\bigl[1 + (T/t_0)\bigr]$,
from the $t = 0$ case of Eq. \eqref{eq:expqt}\big].
%
Using Eq. \eqref{eq:alpha_invt1} in
Eqs. \eqref{eq:ER_sbin} and \eqref{eq:EA_sbin}
yields
\begin{align}
  \Err_R(T)
  &= \frac{ \Err(0) \, t_0^2 } { (T + t_0)^2 }
  ,
  \notag
  %\label{eq:ER_sbin1}
  \\
  \Err_A(T)
  &= \frac{ \Gamma \, T } { (T + t_0)^2 }
  ,
  \label{eq:EA_sbin1}
\end{align}
%
and the total has a minimum of
\begin{equation}
  \Err(T)
  =
  \frac{ \Gamma } { T + t_0 }
  ,
  \label{eq:Emin_sbin}
\end{equation}
%
which is reached at $t_0 = \Gamma /\Err(0)$,
or if we assume Eq. \eqref{eq:error_equil_sbin} for $\Err(0)$,
%
\begin{equation}
  t_0 = \frac{ 2 } { a_0 } = \frac{1}{ \alpha(t = 0) }
  .
  \label{eq:t0_sbin}
\end{equation}
%
The minimum error can be related to
the expected histogram fluctuation\cite{zhou2005, zhou2008}
(cf. Appendix \ref{sec:hfluc}).

@@

We now turn to a general multiple-bin updating scheme.
%
By definition, Eq. \eqref{eq:mbin_update},
a visit to bin $j$ in this case results in updates of the bias potential
not only at bin $j$, but also at a few neighboring bins $i$'s.
%
Similar to Eq. \eqref{eq:ut_diffeq},
we can approximate the updating scheme
as a differential equation,
%
\begin{equation}
  \dot u_i(t)
  \approx
  \alpha(t) \,
  \sum_{j=1}^n w_{ij} \, f_j(t)
  =
  \alpha(t) \,
  \sum_{j=1}^n w_{ij} \, \frac{ h_j(t) } { \rho_j }
  .
  \label{eq:ut_diffeq_mbin}
\end{equation}
%
The optimization procedure is similar to
the single-bin case.
%
The key simplification comes from the
projection of the bias potential to the $n$ eigenmodes
of the updating matrix, $\mathbf w$,
formed by the relative magnitudes, $w_{ij}$'s.

\subsubsection{\label{sec:updating-matrix}
Updating matrix and eigenmode decomposition}

The matrix, $\mathbf w$, is subject to several conditions.
%
First, we have a fixed-point condition\cite{bussi2006, dama2014}.
%
To sample the desired distribution,
$\pmb\rho = (\rho_1, \dots, \rho_n)$,
the growth rate of the bias potential,
${\dot u}_i(t)$
in Eq. \eqref{eq:ut_diffeq_mbin},
should be a constant of $i$
if $h_j(t)$ were the same as $\rho_j$,
i.e.
$\sum_{j=1}^n w_{ij}$ should be a constant
to allow the bias potential to grow uniformly
in the asymptotic regime.
%
By a proper overall scaling of $\mathbf w$,
we may write this condition as
%
\begin{equation}
  \sum_{j = 1}^n w_{ij} = 1
  .
\label{eq:w_sumj}
\end{equation}
%
In other words, $(1, \dots, 1)^T$
is a right eigenvector of $\mathbf w$
with eigenvalue $1$.
%
Thus, the transpose $\mathbf w^T$
resembles a transition matrix,
except that some elements can be negative
in our cases.
%
Below we will take advantage of the analogy
and borrow techniques used
in studying transition matrices\cite{vankampen}
(we will only use properties that are unaffected
by the admission of negative elements into $\mathbf w$).

@@

We can further define
a generalized Fourier transform, $\mathcal{F}$,
for variable, $x_i$,
%
\begin{equation}
  {\tilde x}_k
  \equiv \mathcal{F}[x_i]_k
  \equiv \sum_{i = 1}^n \phi_{ki} \, x_i
  ,
  \notag
  %\label{eq:gft_def}
\end{equation}
%
and the inverse transform is
$x_i = \sum_{k = 0}^{n-1} \phi_{ki} \, \tilde{x}_k / \rho_i$.
%
%\begin{equation}
%  x_i = \frac{1}{\rho_i} \sum_{k = 0}^{n-1} \phi_{ki} \, \tilde{x}_k
%  .
%  \notag
%  %\label{eq:gft_inv}
%\end{equation}
%
Using this definition,
we can rewrite Eq. \eqref{eq:ortho1} as
$\mathcal F[1]_k = \delta_{k0}$,
%
and Eq. \eqref{eq:eigenmode0} means that
the bin average
%of any quantity, $x_i$, is the first mode of the Fourier transform:
$\bav{x}= \mathcal F[x_i]_0 = \tilde x_0$,
%
and
\begin{equation}
  \tilde x_{*k}
  = \tilde x_k - \bav{x} \, \delta_{k0}
  =
  \begin{dcases}
    0           & k = 0
    ,
    \\
    \tilde x_k  & k > 0
    .
  \end{dcases}
  \label{eq:xtstar}
\end{equation}
%
Thus, the only effect
of the bin-average deduction on $\tilde x_k$ is to
annihilate the first mode, $\tilde x_0$.
%
It follows Parseval's theorem,
\begin{equation}
  \sum_{k=1}^{n-1} \tilde x_k \, \tilde y_k
  =
  \sum_{k=0}^{n-1} \tilde x_{*k} \, \tilde y_{*k}
  =
  \sum_{i=1}^n x_{*i} \, y_{*i}
  \label{eq:parseval}
\end{equation}

We can now use Eq. \eqref{eq:eig_w} to diagonalize
the multiple-bin updating scheme, Eq. \eqref{eq:mbin_update},
as
%
$$
{\tilde v}_k(t + 1) =
{\tilde v}_k(t) + \alpha(t) \, \lambda_k \,
{\tilde f}_k(t)
,
$$
or in continuous time, we have, for $k > 0$,
%
\begin{equation}
  \dot{\tilde v}_k(t)
  =
  \alpha(t) \, \lambda_k \, {\tilde f}_k(t)
  \approx
  -\alpha(t) \, \lambda_k \,
  \bigl[ {\tilde v}_k(t) - {\tilde \zeta}_k(t) \bigr]
  ,
  \label{eq:vt_diffeq_mbin}
\end{equation}
%
where in the second step,
we have used Eq. \eqref{eq:xtstar} and
the transformed version of Eq. \eqref{eq:sh_ave}
by assuming the linear approximation
as in Sec. \ref{sec:sbin_diffeq}.

Equation \eqref{eq:vt_diffeq_mbin}
shows that the rate of error reduction
is given by $\alpha(t) \, \lambda_k$.
%
Thus, a \emph{stable} updating scheme
permits no negative eigenvalue, $\lambda_k$,
for $k > 0$.
%
With a set of heterogeneous eigenvalues,
the errors of the eigenmodes are reduced at different rates.
%
Thus, the optimization problem on the final error
becomes more complicated and it often
involves a compromise of the individual modes.

@@

By Eq. \eqref{eq:parseval},
we can rewrite the error defined in Eq. \eqref{eq:error_def} as
$\Err(T) = \sum_{k = 1}^{n - 1} \left\langle {\tilde v}_k^2(T) \right\rangle$.
%
%\begin{align}
%  \Err(T)
%  =
%  \sum_{k = 1}^{n - 1}
%    \left\langle
%      {\tilde v}_k^2(T)
%    \right\rangle
%  .
%\notag
%\end{align}
%
As in Eq. \eqref{eq:error_tot},
we can decompose the error as the sum
of the residual error and the asymptotic error, with
\begin{align}
  \Err_R(T)
  &=
  \sum_{k = 1}^{n-1}
    \left\langle
      {\tilde v}_k^2(0)
    \right\rangle \,
    e^{ - 2 \, \lambda_k  \, q(T) }
  ,
  \label{eq:error_res}
  \\
  \Err_A(T)
  &=
  \int_0^T
  \sum_{k = 1}^{n-1}
  \Gamma_k \, \dot \vartheta_k^2\bigl( q(t) \bigr) \, dt
  ,
  \label{eq:error_asym}
\end{align}
%
where
%
\begin{equation}
  \vartheta_k(q') \equiv e^{\lambda_k \, [q' - q(T)]}
  ,
  \label{eq:uk_def}
\end{equation}
%
and $\Gamma_k$ is computed from
the autocorrelation function
of the histogram fluctuation,
$\bigl\langle {\tilde \zeta}_k(0)
\, {\tilde \zeta}_k(t) \bigr\rangle$,
as:
%
\begin{equation}
  \Gamma_k
  = \sum_{t = -\infty}^\infty
  \bigl\langle {\tilde \zeta}_k(0)
            \, {\tilde \zeta}_k(t) \bigr\rangle
  .
  \label{eq:Gamma_sum}
\end{equation}
%
%\begin{equation}
%$\Gamma_k = \int_{-\infty}^\infty \kappa_k(t) \, dt$,
%  \notag
%\label{eq:Gamma_integral}
%\end{equation}
%
Equation \eqref{eq:Gamma_sum} provides a way
of approximately modeling
a general histogram fluctuation, ${\tilde \zeta}_k(t)$,
as an equivalent white noise.
In this way,
the underlying sampling process
affects the error only through the
few numbers, $\Gamma_k$'s.
%
In Appendix \ref{sec:Gamma_measure},
we discuss a way of estimating $\Gamma_k$
from an FDS simulation under constant magnitude.
%

%\subsubsection{\label{sec:eqlerr}
%Initial error
%%from a simulation under a constant updating magnitude
%}

To model the initial error for Eq. \eqref{eq:error_res},
we will assume that
the preliminary adaptive FDS sampling has been
performed under a constant updating magnitude for $T_0$ steps.
%
If the starting bias potential is zero,
one can show that
\begin{equation}
  \left\langle
    {\tilde v}_k^2(0)
  \right\rangle
  = \frac 1 2 \, a_0 \, \Gamma_k \, \lambda_k
  + \epsilon_k
  ,
  \label{eq:xt2_eql1}
\end{equation}
%
where
$\epsilon_k \equiv \tilde u_k^2 \, e^{-2\, \lambda_k \, a_0 \, T_0}$
and if no eigenvalue is zero,
then we can ignore the $\epsilon_k$ term
in the case of $T_0 \to \infty$,
%
\begin{equation}
  \left\langle
    {\tilde v}_k^2(0)
  \right\rangle
  = \frac 1 2 \, a_0 \, \Gamma_k \, \lambda_k
  .
  \label{eq:xt2_eql}
\end{equation}
%
\note{
Consider a long period, $T_0$, under a fixed schedule,
$\alpha(t) = a_0$,
the residual error becomes negligible, and
the $k$th component of the asymptotic error
is given by
%
\begin{align*}
  \left\langle
    {\tilde v}_k^2(T_0)
  \right\rangle
  =
  \int_0^{T_0}
    \Gamma_k \, (\lambda_k \, a_0)^2 \,
      e^{ 2 \, \lambda_k \, a_0 \, (t - T_0) }
    \, dt
  \stackrel{ T_0 \to \infty }
  { =\joinrel=\joinrel=\joinrel= }
  \frac 1 2 \, \Gamma_k \, \lambda_k \, a_0
  .
\notag
%\label{eq:error_eql}
\end{align*}
}
%
